Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-03 talk/dls2014/paper/paper.tex
Add comment

While this GIL prevents any parallelism from occurring, it also
provides some useful guarantees. Since this lock is always acquired
while executing bytecode instructions and it may only be released
in-between such instructions, it provides perfect isolation and
atomicity between multiple threads for a series of
instructions. Additionally, it provides the application with a
sequential consistency model~\cite{lamport79}. Another technology that
can provide the same guarantees is transactional memory (TM).

@@

TM systems can be broadly categorised as hardware based (HTM),
software based (STM), or hybrid systems (HyTM). HTM systems are limited
by hardware constraints~\cite{odaira14,fuad10}, while STM systems have
a lot of overhead~\cite{cascaval08,drago11}. In \cite{wayforward14},
we argue that STM is still the best way forward, especially since it
supports large atomic blocks as a new way for synchronising multiple
threads. There have been several attempts at lowering the overhead
of STM~\cite{warmhoff13,spear09} - sometimes at the cost of scalability.
In this paper, we describe how we manage to lower the overhead of our
STM system so that it can be seen as a viable replacement for the GIL.

@@

We now use these mechanisms to provide isolation for transactions.  We
implement a \emph{Copy-On-Write (COW)} on the level of
pages~\footnote{Conflict detection still occurs on the level of
objects.}. Starting from the initial fully-shared configuration
(figure \ref{fig:Page-Remapping}, (II)), when we need to modify an
object without other threads seeing the changes immediately, we ensure
that all pages belonging to the object are private to our segment.

More precisely, this is done with a write barrier that detects that we are
about to write to an object that we did not record
in the write-set yet.  When this occurs, the slow-path of the write barrier
will also check if the page (or pages) containing the object is still
shared, and if so, privatise it.  This is done by remapping and copying
the page to the thread's own segment. From now on, the translation
of $\%gs{::}SO$ in this particular thread will resolve to a private
version of the object automatically. Note that the $SO$ used to reference
the object does not change during that process.

@@

To provide atomicity for a transaction, we want to make changes globally
visible on commit. We also need to be able to completely abort a
transaction without a trace, like it never happened.
\begin{description}
\item [{Commit:}] If a transaction commits, we synchronise all threads
  so that all of them are waiting in a safe point. In the committing
  transaction, we go through all objects in the write set and check if
  another transaction in a different segment read the same object.
  Conflicts are resolved again by either the committing or the other
  transaction waiting or aborting.\\
  We then push all changes of modified objects in private pages to all
  the pages in other segments, including the sharing-segment (segment
  0).
\item [{Abort:}] On abort the transaction will forget about all the
  changes it has done. All objects in the write set are reset by
  copying their previous version from the sharing-segment into the
  private pages of the aborting transaction.\\
  Re-sharing these pages instead of resetting the changes would not
  only be slower because it involves a system call, it is also very
  likely the page would get privatised again immediately after re-trying
  the transaction.  Since this again involves a full page copy,
  resetting should be faster than re-sharing.
\end{description}

@@

We use coarse-grained locking for the first three benchmarks and no
locking for the last three. This is important because Jython, which
uses fine-grained locking instead of a GIL, is only expected to scale
with the number of threads for the latter group. It is not able to
scale when using coarse-grained locking. STM, however, uses atomic
blocks instead of a single lock to synchronise accesses to the
shared data structures. Since atomic blocks map to transactions,
our STM system may still be able to get a speedup on more than one
thread by running transactions in parallel.

@@

As expected, all interpreters with a GIL do not scale with the number
of threads. They even become slower because of the overhead of
thread-switching and GIL handling (see \cite{beazley10} for a detailed
analysis). We also see Jython scale when we expect it to (\emph{mandelbrot,
raytrace, richards}), and behave similar to the GIL interpreters in the
other cases. The reason again being the coarse grained locking.

PyPy using our STM system (\emph{pypy-stm-nojit}) scales in all
benchmarks to a certain degree. It scales best in the ones where
Jython scales as well and a little less in the others. The reason
for that is that in the former group there are no real, logical
conflicts -- all threads do independent calculations. In the latter
case, the threads work on a common data structure and therefore
create much more conflicts, which limits the scalability.

Looking at the average overhead from switching from GIL to STM, we see
that it is \remi{$\approx 35.5\%$}. The maximum in richards is
\remi{$63\%$}.

\emph{pypy-stm-nojit} beats \emph{pypy-nojit} already on two threads;
however, it never even beats CPython, the reference implementation of
Python. This means that without the JIT, our performance is not
competitive. So we will now look at how well our system works when we
enable the JIT.

\begin{figure}[h]
  \centering
  \includegraphics[width=1\columnwidth]{plots/performance_nojit.pdf}
  \caption{Comparing execution time between interpreters without a JIT\label{fig:performance-nojit}}
\end{figure}

@@

The results are presented in figure \ref{fig:performance-nojit}. We
see that the performance is much less stable. There is certainly more
work required in this area. The slowdown factor for switching from GIL
to STM ranges around \remi{$1-2.4\times$}, and we beat GIL performance
in half of the benchmarks.

We see that generally, the group of non-locked benchmarks scales
best. The other three scale barely or not at all with the number of
threads. The reason for this is likely again the conflicts in the
latter group. Since the JIT accelerates all code but not the STM
overhead, we do more work per transaction. And this increases the
likelihood of conflicts between them and therefore limits scalability
even more than in the no-JIT benchmarks.

Overall PyPy needs the JIT in order for its performance to be
competitive. It would be interesting to see how using our STM system
in CPython would turn out, but it is a lot of work. On its own, our
system scales well, so we hope to also see that with the JIT in the
future.

\begin{figure}[h]
  \centering
  \includegraphics[width=1\columnwidth]{plots/performance.pdf}
  \caption{Comparing execution time between interpreters with JIT\label{fig:performance-jit}}
\end{figure}

@@

Similar STMs:
\begin{itemize}
\item FastLane: \cite{warmhoff13}
\item TML: \cite{spear09}
\item Virtualizing HTM: \cite{rajwar05}
\item Page-based virtualizing HyTM: \cite{chung06}: page-level conflict
  detection, otherwise hardware extensions required; assumes most
  transactions fit HTM capacities (not so true here); COW using page-faults;
  they assume OS-level access to page-tables (maybe not inherent to their
  design); eval on simulator; value-based confl detection;

 (XTM can be
  implemented either in the OS as part of the virtual memory manager or
  between underlying TM systems and the OS, like virtual machines;
  Conflicts for overflowed transactions are tracked at page granularity;
  XTM-e allows conflict detection at cache line granu-
  larity, even for overflowed data in virtual memory)
\item using mmap(): Memory-Mapped Transactions
\item mem-protected conflict detection: \cite{martin09}
\end{itemize}

@@

% \bibitem{dan07}
%   Dan Grossman. 2007. The transactional memory / garbage collection
%   analogy. \emph{In Proceedings of the 22nd annual ACM SIGPLAN
%     conference on Object-oriented programming systems and
%     applications} (OOPSLA '07).

@@

% \bibitem{felber07}
%   Felber, Pascal, et al. "Transactifying applications using an open
%   compiler framework." \emph{TRANSACT}, August (2007): 4-6.

% \bibitem{bill06}
%   Bill McCloskey, Feng Zhou, David Gay, and Eric
%   Brewer. 2006. Autolocker: synchronization inference for atomic
%   sections. \emph{In Conference record of the 33rd ACM SIGPLAN-SIGACT
%   symposium on Principles of programming languages (POPL '06)}. ACM,
%   New York, NY, USA

@@

% \bibitem{leis14}
%   Leis, Viktor, Alfons Kemper, and Thomas Neumann. "Exploiting
%   Hardware Transactional Memory in Main-Memory Databases."
%   \emph{Proc. of ICDE}. 2014.

% \bibitem{biased}
%   Kenneth Russell and David Detlefs. 2006. Eliminating
%   synchronization-related atomic operations with biased locking and
%   bulk rebiasing. \emph{In Proceedings of the 21st annual ACM SIGPLAN
%     conference on Object-oriented programing, systems, languages, and
%     applications} (OOPSLA '06).

>>>>

While this GIL prevents any parallelism from occurring, it also
provides some useful guarantees. Since this lock is always acquired
while executing bytecode instructions and it may only be released
in-between such instructions, it provides perfect isolation and
atomicity between multiple threads for a series of
instructions. Additionally, it provides the application with a
sequential consistency model~\cite{lamport79}. Another technology that
can provide the same guarantees is transactional memory
(TM). \remi{cite our position paper}

@@

TM systems come in\arigo{typo?} can be broadly categorised as hardware based (HTM),
software based (STM), or hybrid systems (HyTM). HTM systems are limited
by hardware constraints~\cite{odaira14,fuad10}, while STM systems have
a lot of overhead~\cite{cascaval08,drago11}. In \cite{wayforward14},
we argue that STM is still the best way forward, especially since it
supports large atomic blocks as a new way for synchronising multiple
threads. There have been several attempts at lowering the overhead
of STM~\cite{warmhoff13,spear09} - sometimes at the cost of scalability.
In this paper, we describe how we manage to lower the overhead of our
STM system so that it can be seen as a viable replacement for the GIL.

@@

We now use these mechanisms to provide isolation for transactions.
Using write barriers, we implement a \emph{Copy-On-Write (COW)} on the
level of pages~\footnote{Conflict detection still occurs on the level
of objects.}. Starting from the initial fully-shared configuration
(figure \ref{fig:Page-Remapping}, (II)), when we need to modify an
object without other threads seeing the changes immediately, we ensure
that all pages belonging to the object are private to our segment.

More precisely, this is done by a write barrier that detects that we are
about to write to an old (pre-transaction) object that we did not record
in the write-set yet.  When this occurs, the slow-path of the write barrier
will also check if the page (or pages) containing the object is still
shared, and if so, privatise it.  This is done by remapping and copying
the page to the thread's own segment. From now on, the translation
of $\%gs{::}SO$ in this particular thread will resolve to a private
version of the object automatically. Note that the $SO$ used to reference
the object does not change during that process.

@@

To provide atomicity for a transaction, we want to make changes globally
visible on commit. We also need to be able to completely abort a
transaction without a trace, like it never happened.
\begin{description}
\item [{Commit:}] If a transaction commits, we synchronise all threads
  so that all of them are waiting in a safe point.\arigo{Explain why it's
  not completely horribly slow here} In the committing
  transaction, we go through all objects in the write set and check if
  another transaction in a different segment read the same object.
  Conflicts are resolved again by either the committing or the other
  transaction waiting or aborting.\\
  We then push all changes of modified objects in private pages to all
  the pages in other segments, including the sharing-segment (segment
  0).
\item [{Abort:}] On abort the transaction will forget about all the
  changes it has done. All objects in the write set are reset by
  copying their previous version from the sharing-segment into the
  private pages of the aborting transaction.\\
  Re-sharing these pages instead of resetting the changes would not
  only be slower because it involves a system call, it is also very
  likely the page would get privatised again immediately after re-trying
  the transaction.  Since this again involves a full page copy,
  resetting should be faster than re-sharing.
\end{description}

@@

We use coarse-grained locking for the first three benchmarks and no
locking for the last three. This is important because Jython, which
uses fine-grained locking instead of a GIL, is only expected to scale
with the number of threads for the latter group. It is not able to
scale when using coarse-grained locking. STM, however, uses atomic
blocks instead, so it may still be able to scale since they are
implemented as simple transactions.

@@

As expected, all interpreters with a GIL do not scale with the number
of threads. They even become slower because of the overhead of
thread-switching and GIL handling (see \cite{beazley10} for a detailed
analysis). We also see Jython scale when we expect it to (mandelbrot,
raytrace, richards), and behave similar to the GIL interpreters in the
other cases.

PyPy using our STM system (pypy-stm-nojit) scales in all benchmarks to
a certain degree. We see that the average overhead from switching from
GIL to STM is \remi{$35.5\%$}, the maximum in richards is
\remi{$63\%$}. pypy-stm-nojit beats pypy-nojit already on two threads;
however, it never even beats CPython, the reference implementation of
Python. This means that without the JIT, our performance is not
competitive. We now look at how well our system works when we enable
the JIT.

\begin{figure}[h]
  \centering
  \includegraphics[width=1\columnwidth]{plots/performance_nojit.pdf}
  \caption{Comparing runtime between interpreters without a JIT\label{fig:performance-nojit}}
\end{figure}

@@

The results are presented in figure \ref{fig:performance-nojit}. We
see that the performance is much less stable. There is certainly more
work required in this area. In general, we see that the group of
non-locked benchmarks certainly scales best. The other three scale
barely or not at all with the number of threads. The slowdown factor
from GIL to STM ranges around \remi{$1-2.4\times$} and we beat GIL
performance in half of the benchmarks.

\remi{Reason for bad scaling: acceleration of code that produces
conflicts $-->$ more iterations $-->$ more conflicts. The overhead
doesn't get accelerated by the JIT.}

\begin{figure}[h]
  \centering
  \includegraphics[width=1\columnwidth]{plots/performance.pdf}
  \caption{Comparing runtime between interpreters with JIT\label{fig:performance-jit}}
\end{figure}

Overall PyPy needs the JIT in order for its performance to be
competitive.  It would be interesting to see how using our STM system
in CPython would perform, but it is a lot of work. On its own, our
system scales well so we hope to also see that with the JIT in the
future.

@@

Similar STMs:
\begin{itemize}
\item FastLane: \cite{warmhoff13}
\item TML: \cite{spear09}
\item Virtualizing HTM: \cite{rajwar05}
\item Page-based virtualizing HyTM: \cite{chung06} (XTM can be
  implemented either in the OS as part of the virtual memory manager or
  between underlying TM systems and the OS, like virtual machines;
  Conflicts for overflowed transactions are tracked at page granularity;
  XTM-e allows conflict detection at cache line granu-
  larity, even for overflowed data in virtual memory)
\item using mmap(): Memory-Mapped Transactions
\item mem-protected conflict detection: \cite{martin09}
\end{itemize}

@@

\bibitem{dan07}
  Dan Grossman. 2007. The transactional memory / garbage collection
  analogy. \emph{In Proceedings of the 22nd annual ACM SIGPLAN
    conference on Object-oriented programming systems and
    applications} (OOPSLA '07).

@@

\bibitem{felber07}
  Felber, Pascal, et al. "Transactifying applications using an open
  compiler framework." \emph{TRANSACT}, August (2007): 4-6.

\bibitem{bill06}
  Bill McCloskey, Feng Zhou, David Gay, and Eric
  Brewer. 2006. Autolocker: synchronization inference for atomic
  sections. \emph{In Conference record of the 33rd ACM SIGPLAN-SIGACT
  symposium on Principles of programming languages (POPL '06)}. ACM,
  New York, NY, USA

@@

\bibitem{leis14}
  Leis, Viktor, Alfons Kemper, and Thomas Neumann. "Exploiting
  Hardware Transactional Memory in Main-Memory Databases."
  \emph{Proc. of ICDE}. 2014.

\bibitem{biased}
  Kenneth Russell and David Detlefs. 2006. Eliminating
  synchronization-related atomic operations with biased locking and
  bulk rebiasing. \emph{In Proceedings of the 21st annual ACM SIGPLAN
    conference on Object-oriented programing, systems, languages, and
    applications} (OOPSLA '06).
